---
title: "modelo SUPERVIVENCIA EPIFITAS"
author: "Edith"
format: html
editor: visual
---

## Relocación Epifitas a la puna

Paquetes

```{r}
#| warning: false
#| message: false

library(car)
library(dplyr)
library(performance)
library(DHARMa)
library(emmeans)
```

Base de datos y preparacion de las variables para los modelos.

Quite los individuos despues de muertos

```{r}

# NUNCA NECESITAS TO DO "SETWD" CUANDO USAS PROYECTOS
# Borra estas tres líneas
#setwd("C:/Users/edith clemente arena/Desktop/EPIFITAS -GRADIENTE/treeline-epiphytes/treeline-epiphytes/Doble_filtro/Data")

expdata <- read.csv("../Data/relocadata.csv", 
                    header = TRUE, sep = ";", 
                    stringsAsFactors = FALSE)
str(expdata)      # Estructura de la base de datos

##Preparar tabla de datos 
data_model <- expdata[, c("epi_ID", "dead_alive", "relocation", "time", "branch_ID")]
data_model$time <- factor(data_model$time)
#recodificar variables 
data_model$survive <- ifelse(data_model$dead_alive == "alive", 1, 0)
data_model$relocation[data_model$relocation == "true"] <- "puna"
data_model$relocation[data_model$relocation == "false"] <- "bosque"
data_model$relocation <- factor(data_model$relocation)
data_model$branch_ID <- factor(data_model$branch_ID)
data_model$epi_ID <- factor(data_model$epi_ID)

#QUITA LOS MUERTOS
data_model <- data_model %>%
  arrange(epi_ID, time) %>%
  group_by(epi_ID) %>%
  mutate(dead_seen_before = lag(cumany(dead_alive == "dead"), default = FALSE)) %>%
  filter(!dead_seen_before) %>%
  ungroup() %>%
  dplyr::select(-dead_seen_before)
#Verificacion
data_model %>%
  arrange(epi_ID, time) %>%
    dplyr::select(epi_ID, time, dead_alive, survive) %>%
  print(n = 150)
table(data_model$epi_ID)#cuenta cuantas observaciones por individuo
table(data_model$time, data_model$survive)#ver los vivos o muertos por tiempos
#write.csv(data_model, "data_model.csv", row.names = FALSE)
#Cambiar el nombre de la rama 
data_model <- data_model %>%
  mutate(branch_ID = case_when(
    branch_ID == "B01" ~ "PU01",
    branch_ID == "B02" ~ "PU02",
    branch_ID == "B03" ~ "PU03",
    branch_ID == "B04" ~ "PU04",
    branch_ID == "B05" ~ "PU05",
    branch_ID == "B06" ~ "BO06",
    branch_ID == "B07" ~ "BO07",
    branch_ID == "B08" ~ "BO08",
    branch_ID == "B09" ~ "BO09",
    branch_ID == "B10" ~ "BO10"
  )) %>%
  mutate(branch_ID = factor(branch_ID))
```

Nota:

Todas las muertes se dieron en la puna y ninguna en el bosque, ¿produce separación?

los datos en los diferentes tiempos se van reduciendo porque solo quedan los vivos.

## Pregunta

¿Disminuye la supervivencia de las epífitas cuando son trasladadas del treeline a condiciones de puna?

## Modelos

#### Modelo interacción: si la relocacion afecta la sobrevivencia y cambia con el tiempo

No produce sobredispercion,

```{r}
model_glm_int <- glm(
  survive ~ relocation*time,
  family = binomial,
  data = data_model
)
summary(model_glm_int)
Anova(model_glm_int)
#SOBREDISPERSION
check_overdispersion(model_glm_int)
#DIAGNOSTICO con residuos simulados
sim_res <- simulateResiduals(model_glm_int)
plot(sim_res)
#Revisar los supuestos 
dev.new()
check_model(model_glm_int)

```

Nota: se observa la colinearidad entre la interaccion y sus partes pero es porque son parte

### Modelo interaccion con rama fijo:

Provoca sobreparametrizacion y produce overfitting

```{r}
model_glm_intb <- glm(
  survive ~ relocation*time + branch_ID,
  family = binomial,
  data = data_model
)
summary(model_glm_intb)
Anova(model_glm_intb)
#SOBREDISPERSION
check_overdispersion(model_glm_intb)#sin sobredispersion
##DIAGNOSTICO GENERAL
sim_resb <- simulateResiduals(model_glm_intb)#El modelo NO está capturando bien la relación
plot(sim_resb)
dev.new()
check_model(model_glm_intb)

```

Solo tiene efecto significativo el tiempo, en la relocacion el modelo no puede evaluarlo correctamente probablemente por la sobreparametrizacion y la separacion perfecta de puna y bosque (todos vivos) .

Al igual que el otro modelo posee colinearidad.

#### NOTA

Ambos modelos poseen coeficientes bajisimos, errores estandar gigantes y p_valores raros, será porque el efecto es muy fuerte?













Glenda's
Do forest and puna differ in survival over censuses?


```{r}
data_surv <- subset(data_model, time != "T1")


#Best fix: Firth logistic regression

#Using brglm2:

#install.packages("brglm2")
library(brglm2)

model_br <- glm(
  survive ~ relocation * time,
  data = data_surv,
  family = binomial,
  method = "brglmFit"
)

summary(model_br)
#Anova(model_br)

#This is often the best replacement for your current glm() when there is separation.

#✅ Correct way to test effects in brglm2

#Fit nested models:

m_add <- glm(
  survive ~ relocation + time,
  data = data_surv,
  family = binomial,
  method = "brglmFit"
)

m_rel <- glm(
  survive ~ relocation,
  data = data_surv,
  family = binomial,
  method = "brglmFit"
)

m_time <- glm(
  survive ~ time,
  data = data_surv,
  family = binomial,
  method = "brglmFit"
)

m_null <- glm(
  survive ~ 1,
  data = data_surv,
  family = binomial,
  method = "brglmFit"
)

#Then compare them.

#1) Test the interaction
anova(m_add, model_br, test = "LRT")

#This tests:

#Does relocation × time improve the model?

#2) Test relocation effect
anova(m_time, m_add, test = "LRT")

#This tests:

#Does adding relocation improve the model beyond time alone?

#3) Test time effect
anova(m_rel, m_add, test = "LRT")

#This tests:

#Does adding time improve the model beyond relocation alone?
```



```{r}

#🏆 Best next step: predict survival probabilities

#Now that you have the right model, let’s get the predicted survival probabilities by treatment and census.

#✅ Recommended prediction workflow
#Option A (best): use emmeans
library(emmeans)

pred_surv <- emmeans(
  model_br,
  ~ relocation * time,
  type = "response"
)

pred_df <- as.data.frame(pred_surv)
pred_df

#This gives you:

#predicted probability of survival (prob)
#confidence intervals
#for each:
#bosque / puna
#T2 / T3 / T4
#📊 Plot predicted survival probabilities
library(ggplot2)

ggplot(pred_df, aes(x = time, y = prob, color = relocation, group = relocation)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.08) +
  ylim(0, 1) +
  labs(
    x = "Census",
    y = "Predicted survival probability",
    color = "Relocation"
  ) +
  theme_classic()
#⭐ Better figure: add the observed proportions too

#This is often the best ecological figure because it shows:

#points = observed survival proportions
#lines = model predictions
#Step 1 — Calculate observed survival by group
obs_df <- data_surv %>%
  group_by(relocation, time) %>%
  summarise(
    obs_surv = mean(survive),
    n = n(),
    .groups = "drop"
  )
#Step 2 — Plot observed + predicted together
ggplot() +
  geom_point(data = obs_df,
             aes(x = time, y = obs_surv, color = relocation),
             size = 3, alpha = 0.8) +
  geom_line(data = pred_df,
            aes(x = time, y = prob, color = relocation, group = relocation),
            linewidth = 1) +
  geom_errorbar(data = pred_df,
                aes(x = time, y = prob, ymin = asymp.LCL, ymax = asymp.UCL, color = relocation),
                width = 0.08) +
  ylim(0, 1) +
  labs(
    x = "Census",
    y = "Survival probability",
    color = "Relocation"
  ) +
  theme_classic()

#This is probably the best figure for your paper/presentation.
```




```{r}
#⚙️ Step 1 — Prepare the data

#Make sure time is ordered correctly:

library(dplyr)

data_model <- data_model %>%
  mutate(time = factor(time, levels = c("T1", "T2", "T3", "T4")))

#⚙️ Step 2 — Create numeric time

#We need numeric time for survival models:

data_model <- data_model %>%
  mutate(time_num = as.numeric(time))

#So:

T1 = 1
T2 = 2
T3 = 3
T4 = 4
#⚙️ Step 3 — Collapse to one row per individual

#This is the key step:

surv_data <- data_model %>%
  group_by(epi_ID) %>%
  summarise(
    relocation = first(relocation),

    # Did the individual ever die?
    event = ifelse(any(survive == 0), 1, 0),

    # Time of death OR last observed time
    time_to_event = ifelse(
      any(survive == 0),
      min(time_num[survive == 0]),   # first time it died
      max(time_num)                  # last time observed alive
    )
  ) %>%
  ungroup()
#🔍 Check the result
table(surv_data$event)
summary(surv_data$time_to_event)
head(surv_data)


#Firth-corrected Cox model
#install.packages("coxphf")
library(coxphf)
library(survival)

cox_firth <- coxphf(
  Surv(time_to_event, event) ~ relocation,
  data = surv_data
)

summary(cox_firth)
```

Survival differed strongly between relocation treatments. Standard Cox regression showed evidence of quasi/complete separation, indicating extremely uneven mortality between groups, so treatment effects should be interpreted using penalized Cox regression and Kaplan–Meier/log-rank comparisons rather than ordinary maximum-likelihood Cox estimates.


```{r}

#Best compact diagnostic script for your exact use case

#If you want a practical “run this and inspect” block, use this:

library(brglm2)
library(car)
library(DHARMa)
library(pROC)

model_br <- glm(
  survive ~ relocation * time,
  data = data_surv,
  family = binomial,
  method = "brglmFit"
)

# 1. Summary + ORs
summary(model_br)
confint(model_br)
exp(cbind(OR = coef(model_br), confint(model_br)))

# 2. Sparse cells / separation structure
with(data_surv, table(relocation, time, survive))

# 3. Residuals
data_surv$pearson_resid  <- residuals(model_br, type = "pearson")
data_surv$deviance_resid <- residuals(model_br, type = "deviance")
data_surv$fitted_prob    <- fitted(model_br)

plot(data_surv$fitted_prob, data_surv$deviance_resid,
     xlab = "Fitted probability",
     ylab = "Deviance residuals")
abline(h = 0, lty = 2, col = "red")

# 4. Influence
data_surv$hat   <- hatvalues(model_br)
data_surv$cooks <- cooks.distance(model_br)

plot(data_surv$hat, type = "h", main = "Leverage")
abline(h = 2 * mean(data_surv$hat), col = "red", lty = 2)

plot(data_surv$cooks, type = "h", main = "Cook's distance")
abline(h = 4 / nrow(data_surv), col = "red", lty = 2)

subset(data_surv, abs(deviance_resid) > 2 | cooks > 4 / nrow(data_surv))

# 5. Collinearity
vif(model_br)

# 6. Simulated residuals
sim_res <- simulateResiduals(model_br, n = 1000)
plot(sim_res)
testDispersion(sim_res)
testOutliers(sim_res)

# 7. ROC / AUC
roc_obj <- roc(data_surv$survive, fitted(model_br))
plot(roc_obj)
auc(roc_obj)


```

How to write this up in a paper / thesis

A clean write-up would be:

Because the binomial GLM showed evidence of separation, we refit the model using bias-reduced logistic regression (brglm2, method = "brglmFit"). Model diagnostics included inspection of deviance residuals, leverage, Cook’s distance, variance inflation factors, and simulated residuals. Sparse outcome combinations were also examined directly using contingency tables. No severe residual misfit or undue influence was observed / [or describe what you found].

Bottom line

For Firth / bias-reduced logistic regression, the most useful diagnostics are:

cross-tabulation of predictors vs outcome (to understand separation),
deviance residuals,
leverage / Cook’s distance,
VIF,
DHARMa simulated residuals,
and AUC/calibration if prediction matters.

If you want, paste your summary(model_br) output (and whether time is continuous or categorical), and I can tell you exactly which diagnostics matter most for your model and how to interpret them.